---
title: "`PTCMGH`: Disease Free Survival Analysis"
author: '[Francisco Javier Rubio](https://sites.google.com/site/fjavierrubio67/)'
#date: "`r format(Sys.Date(), "%B %d, %Y")`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

CALGB 40101 (NCT00041119) is a Phase III trial finding that 4 cycles of chemotherapy are non-inferior to 6 cycles for early-stage breast cancer (0-3 positive nodes).


The following R code illustrates how to analyse disease free survival in this data set using the `PTCMGH` R package.


# Data preparation

```{r}

rm(list=ls())

#library(devtools)
#install_github("FJRubio67/HazReg")
library(HazReg)
#library(devtools)
#install_github("FJRubio67/PTCMGH")
library(PTCMGH)
library(survival)
library(knitr)
library(haven)
library(dplyr)

# Read the raw data set
df <- read_sas("all_finalb.sas7bdat")
head(df)

# Data
df_raw <- read_sas("all_finalb.sas7bdat")

df <- df_raw %>%
  # Restrict to eligible patients only
  filter(elig == 2) %>%
  mutate(
    # --- Outcome variables ---
    os_time   = survmos,
    os_event  = survstat,       # 0=alive, 1=dead  
    dfs_time  = dfsmos,
    dfs_event = dfsstat,        # 0=no event, 1=event  
    
    # --- Treatment (primary exposures) ---
    agent  = factor(agent,  levels = c(0,1), labels = c("CA", "T")),
    length = factor(length, levels = c(0,1), labels = c("4 cycles", "6 cycles")),
    
    # --- Stratification / baseline covariates ---
    menopause = factor(stra1, levels = c(1,2), labels = c("Pre", "Post")),
    receptor  = factor(stra2, levels = c(1,2), labels = c("Pos/Unk", "Neg")),
    her2      = factor(stra3, levels = c(1,2,3), labels = c("Positive","Negative","Unknown")),
    
    # --- Tumour characteristics ---
    er_status   = factor(OH003, levels = c(1,2), labels = c("Negative","Positive")),
    pgr_status  = factor(OH004, levels = c(1,2), labels = c("Negative","Positive")),
    grade       = factor(OH005, levels = c(1,2,3), labels = c("Low","Intermediate","High")),
    tumor_size  = factor(tsize, levels = c(1,2,3), labels = c("<2cm","2-5cm",">5cm")),
    num_pos_nodes = num_pos_nodes,   # keep as-is or convert to numeric (note ">3" category)
    
    # --- Demographics ---
    age_cat = factor(agecat, levels = 1:6,
                     labels = c("20-29","30-39","40-49","50-59","60-69","70+")),
    race    = factor(RACE_ID, levels = c(1,3,4,5,99),
                     labels = c("White","Black","Asian","Other","Unknown")),
    ethnicity = factor(ETHNIC_ID, levels = c(1,2,9),
                       labels = c("Hispanic","Non-Hispanic","Unknown")),
    
    # --- Surgery / staging ---
    surgery_type = factor(OH028, levels = c(1,2), labels = c("Lumpectomy","Mastectomy")),
    sentinel_bx  = factor(OH032, levels = c(1,2), labels = c("No","Yes")),
    axillary_diss = factor(OH037, levels = c(1,2), labels = c("No","Yes")),
    
    # --- Amendment (important! 6-cycle arms were later closed) ---
    preamend = factor(preamend, levels = c(0,1), labels = c("Post-amend","Pre-amend"))
  )

dim(df)



# Complete cases for variables of interest
dfc <- data.frame(agent = df$agent, length = df$length, menopause = df$menopause,
                  receptor = df$receptor, tumor_size = df$tumor_size, grade = df$grade,
                  num_pos_nodes = as.numeric(df$num_pos_nodes), age_cat = df$age_cat,
                  indrx = df$indrx, her2 = df$her2,
                  dfs_time = df$dfs_time/12, dfs_event = df$dfs_event)

dfc <- dfc[complete.cases(dfc),]
dim(dfc)


# Build the design matrix from the Cox model formula
des0 <- model.matrix(
  ~ factor(agent) + factor(length) + factor(menopause) +
    factor(receptor) + factor(her2) + factor(tumor_size) +
    factor(grade),
  data = dfc
)

# View the first few rows
head(des0)
dim(des0)
nc <- nrow(dfc)


# Survival times
times <- dfc$dfs_time
# Status
status <- dfc$dfs_event

# Kaplan-Meier estimator for the overall survival times
km <- survfit(Surv(times, status) ~ 1)
plot(km$time, km$surv, type = "l", col = "black", lwd = 2, lty = 1, 
     ylim = c(0,1), xlab = "Time", ylab = "Survival")


# Kaplan-Meier estimator for the overall survival times: T (0) vs CA (1)
ind_t <- which(dfc$agent == "T")
ind_ca <- which(dfc$agent == "CA")

km_t <- survfit(Surv(times[ind_t], status[ind_t]) ~ 1)
km_ca <- survfit(Surv(times[ind_ca], status[ind_ca]) ~ 1)

plot(km$time, km$surv, type = "l", col = "black", lwd = 2, lty = 1, 
     ylim = c(0,1), xlab = "Time", ylab = "Survival")
points(km_t$time, km_t$surv, type = "l", col = "red", lwd = 2, lty = 1, 
     ylim = c(0,1), xlab = "Time", ylab = "Survival")
points(km_ca$time, km_ca$surv, type = "l", col = "green", lwd = 2, lty = 1, 
     ylim = c(0,1), xlab = "Time", ylab = "Survival")
legend("bottomright", legend = c("Population","T","CA"), col = c("black","red","green"), 
       lwd = c(2,2,2))
```

# Model fit

## Baseline

```{r}

# Model fitting
OPT <- PTCMMLE(init = c(0,0,0),
                    times = times,
                    status = status,
                    hstr = "baseline",
                    dist = "LogNormal",
                    des_theta = NULL,
                    des_t = NULL,
                    des_h = NULL,
                    des = NULL,
                    method = "nlminb", 
                    maxit = 10000) 

# MLE
MLE <- c(OPT$OPT$par[1],exp(OPT$OPT$par[2]),exp(OPT$OPT$par[3]))

# Fitted survival function 
spt <- Vectorize(function(t) exp(- MLE[3]*(1-exp(-chlnorm(t,MLE[1],MLE[2])))) )

plot(km$time, km$surv, type = "l", col = "black", lwd = 2, lty = 1, 
     ylim = c(0,1), xlab = "Time", ylab = "Survival")
curve(spt,0,10.5, col = "red", add= T) 
legend("bottomright", legend = c("KM","Baseline"), col = c("black","red"), 
       lwd = c(2,2))
```

## PH 

```{r}
# Model fitting
OPT_PH <- PTCMMLE(init = rep(0,2*ncol(des0)+1),
               times = times,
               status = status,
               hstr = "PH",
               dist = "LogNormal",
               des_theta = des0,
               des_t = NULL,
               des_h = NULL,
               des = des0[,-1],
               method = "nlminb", 
               maxit = 10000) 

# MLE
MLE_PH <- c(OPT_PH$OPT$par[1],exp(OPT_PH$OPT$par[2]),OPT_PH$OPT$par[-c(1:2)])

# Population survival function 
spt_ph <- Vectorize(function(t){
  
  theta_i <- as.vector(exp(des0 %*% MLE_PH[3:(2+ncol(des0))]))
  F_i <- 1 - exp(-chlnorm(t,MLE_PH[1],MLE_PH[2])*exp(des0[,-1]%*%MLE_PH[(3+ncol(des0)):(2*ncol(des0)+1)]))
  survs <- exp(-theta_i*F_i)
  return(mean(survs))
  }) 

plot(km$time, km$surv, type = "l", col = "black", lwd = 2, lty = 1, 
     ylim = c(0,1), xlab = "Time", ylab = "Survival")
curve(spt_ph,0,10.5, col = "red", add= T)  
legend("bottomright", legend = c("KM","PH"), col = c("black","red"), 
       lwd = c(2,2))
```

## AFT 


```{r}
# Model fitting
OPT_AFT <- PTCMMLE(init = rep(0,2*ncol(des0)-1+2),
                  times = times,
                  status = status,
                  hstr = "AFT",
                  dist = "LogNormal",
                  des_theta = des0,
                  des_t = NULL,
                  des_h = NULL,
                  des = des0[,-1],
                  method = "nlminb", 
                  maxit = 10000) 

# MLE
MLE_AFT <- c(OPT_AFT$OPT$par[1],exp(OPT_AFT$OPT$par[2]),OPT_AFT$OPT$par[-c(1:2)])

# Population survival function
spt_aft <- Vectorize(function(t){
  
  theta_i <- as.vector(exp(des0 %*% MLE_AFT[3:(2+ncol(des0))]))
  F_i <- 1 - exp(-chlnorm(as.vector(t*exp(des0[,-1]%*%MLE_AFT[(3+ncol(des0)):(2*ncol(des0)+1)])),MLE_AFT[1],MLE_AFT[2]))
  survs <- exp(-theta_i*F_i)
  return(mean(survs))
}) 


plot(km$time, km$surv, type = "l", col = "black", lwd = 2, lty = 1, 
     ylim = c(0,1), xlab = "Time", ylab = "Survival")
curve(spt_aft,0,10.5, col = "red", add= T)  
legend("bottomright", legend = c("KM","AFT"), col = c("black","red"), 
       lwd = c(2,2))
```


## General Hazards

```{r}
# Model fitting
OPT_GH <- PTCMMLE(init = rep(0,3*ncol(des0)),
                   times = times,
                   status = status,
                   hstr = "GH",
                   dist = "LogNormal",
                   des_theta = des0,
                   des_t = des0[,-1],
                   des_h = des0[,-1],
                   des = NULL,
                   method = "nlminb", 
                   maxit = 10000) 

# MLE
MLE_GH <- c(OPT_GH$OPT$par[1],exp(OPT_GH$OPT$par[2]),OPT_GH$OPT$par[-c(1:2)])


# Population survival function
spt_gh <- Vectorize(function(t){
  
  theta_i <- as.vector(exp(des0 %*% MLE_GH[3:(2+ncol(des0))]))
  F_i <- 1 - exp(-chlnorm(as.vector(t*as.vector(exp(des0[,-1]%*%MLE_GH[(3+ncol(des0)):(2*ncol(des0)+1)]))),MLE_GH[1],MLE_GH[2])*
                   as.vector(exp(des0[,-1]%*%MLE_GH[(2*ncol(des0)+2):(3*ncol(des0))] - des0[,-1]%*%MLE_GH[(3+ncol(des0)):(2*ncol(des0)+1)])))
  survs <- exp(-theta_i*F_i)
  return(mean(survs))
}) 

plot(km$time, km$surv, type = "l", col = "black", lwd = 2, lty = 1, 
     ylim = c(0,1), xlab = "Time", ylab = "Survival")
curve(spt_gh,0,10.5, col = "red", add= T)  
legend("bottomright", legend = c("KM","GH"), col = c("black","red"), 
       lwd = c(2,2))
```

## Model comparison

```{r}
# Akaike information criterion
AIC <- 2*OPT$OPT$objective + 2*length(OPT$OPT$par)
AIC_PH <- 2*OPT_PH$OPT$objective + 2*length(OPT_PH$OPT$par)
AIC_AFT <- 2*OPT_AFT$OPT$objective + 2*length(OPT_AFT$OPT$par)
AIC_GH <- 2*OPT_GH$OPT$objective + 2*length(OPT_GH$OPT$par)

# AICs: baseline, PH, AFT, AH, GH
AICs <- c(AIC, AIC_PH, AIC_AFT, AIC_GH)

AICs

which.min(AICs)

```

# Estimands: PH model

## Incidence

### Difference in cure probabilities

```{r}
# x: covariate values
# index: index of x corresponding to treatment (agent in this case)
diff_CP <- function(x, index){
  x0 <- as.matrix(x) 
  x0[,index] = 0
  x1 <- as.matrix(x)
  x1[,index] = 1
  theta_i0 <- as.vector(exp(x0 %*% MLE_PH[3:(2+ncol(x0))]))
  theta_i1 <- as.vector(exp(x1 %*% MLE_PH[3:(2+ncol(x1))]))
  
  dif_theta <- exp(-theta_i1) - exp(-theta_i0)
  return(as.vector(dif_theta))
}

DCP0 <- diff_CP(des0,2)

hist(DCP0, main = "Difference in cure probabilities",
     xlab = "DCP", probability = TRUE)

plot(sort(DCP0), main = "Difference in cure probabilities")

# Average treatment effect (ATE): Cure probabilities
ATE_CP <- mean(DCP0)
ATE_CP

summary(DCP0)
```

## Difference in latent failure distributions

```{r}
# x: covariate values
# index: index of x corresponding to treatment (agent in this case)
diff_LFD <- function(t, x, index){
  x0 <- as.matrix(x) 
  x0[,index] = 0
  x1 <- as.matrix(x)
  x1[,index] = 1
  F_i0 <- 1 - exp(-chlnorm(t,MLE_PH[1],MLE_PH[2])*exp(x0[,-1]%*%MLE_PH[(3+ncol(x0)):(2*ncol(x0)+1)]))
  F_i1 <- 1 - exp(-chlnorm(t,MLE_PH[1],MLE_PH[2])*exp(x1[,-1]%*%MLE_PH[(3+ncol(x1)):(2*ncol(x1)+1)]))
  
  dif_lfd <- F_i1 - F_i0
  return(as.vector(dif_lfd))
}

ts <- seq(0,10.5, by = 0.05)

DLFD0 <- matrix(0, nrow = nc, ncol = length(ts))
for(i in 1:length(ts)) DLFD0[,i] <- diff_LFD(ts[i], des0, 2)

# Average treatment effect (ATE): latent failure distributions
plot(colMeans(DLFD0), main = "Difference in latent failure distributions",
     xlab = "time", ylab = "DLFD")

```


# References
